### 2016-07-12 capability-machine.tex
added a new symbolic spec. for malloc

In the JS paper, the ticket dispenser program dereferences the counter and returns the value. It does not seem like a similar thing would work here as the adversary can do even more. In particular, the adversary can execute a $\halt$ instruction that will cause the machine to halt (successfully). The adversary could also choose never to return to the return address that we set up for him.
\begin{lemma}[Ticket dispenser test]\dominique{I would prefer more precise, symbolic statements of assumptions and less prose.}
\label{lem:tckt-disp}
 Given a configuration $\Phi$, if
 \begin{description}
 \item[$\boldsymbol \memheap$] contains the ticket dispenser test program starting at the address labelled $\initb$, 
 \item[$\boldsymbol \memheap$] contains the adversary code starting at label $\advb$. The adversary is assumed to have no capabilities on the heap, but it has access to the ambient $\malloc$ capability! \lau{maybe mention that malloc is also on the heap}
 \item[$\boldsymbol \memheap$] contains the malloc code.
 \item[Nothing else is allocated in $\boldsymbol \memheap$] and all the above programs are disjoint.
 \item[$\boldsymbol{\adv}$ and $\boldsymbol{ \codelabel{malloc}}$] capabilities are ambiently available at all times.
 \item[$\boldsymbol{ \memreg(\pcreg)}$] is $(\rwx,\initb,\inite,\initb)$, which is the only capability that grants any access to this part of the heap,
 \item[$\boldsymbol{ \memreg(\rio)}$] is $(\readwrite,0,0,0)$, which also is the only capability that grants access to this part of the heap, 
 \item[the remaining registers] all contains 0,
 \end{description}
 and $\Phi \rightarrow^* (\halted,\Phi')$, then the I/O address (that is $\memheap[\Phi'](0)$) contains either $-1$ or a number that is $\geq 0$ and even.
\end{lemma}
In the initial configuration the $\pcreg$ register is assumed to have an execute and write permission for the area of the heap where the program resides.

The ticket dispenser test program uses $\malloc$. To be able to reason about programs that use $\malloc$, we assume that it follows the following specification. If you provide it with one argument, namely the size of the allocation you wish and a return capability, then at some point $\pcreg$ will contain an executable capability based on this return capability and the return register will contain a $\rwx$ capability for the freshly allocated memory. 
\begin{specification}[Malloc]\dominique{can you state this more systematically, with less prose, more explicit assumptions and conclusion?}
Take $W_{\malloc}$ the world with exactly one island, namely the unspecified island $\iota_{\malloc}$ that governs the internal state of $\malloc$.  If we invoke $\malloc$ in a memory $h = \hsframe \uplus \hsfoot$
%\lau{Is $\hsframe$ supposed to be the rest of memory, so $h$ is something we can execute in? Is $\malloc$ allowed to allocate more memory, i.e., how does $\hsfoot$ relate to $\hsfoot'$?} A: The frame contains anything that has been allocated and the footprint contains the rest. That is the footprint contains mallocs internal state as well as the free memory.
% Notice: the current worlds do not support malloc allocating memory as the invariants cannot evolve.
that is the disjoint sum of a frame part and a footprint part and the latter satisfies this world up to steps $n$, i.e., $\heapSat[\hsfoot]{n}{W_{\malloc}}$. Then $\malloc$ will successfully terminate and return a result in a heap $h' = \hsframe \uplus \hsfoot'$ that is the disjoint sum of the same unmodified frame part and a potentially new footprint part, such that there is a future world $W'$ of $W_{\malloc}$ consisting of exactly two islands, where the first is an instance of $\iota_{\malloc}$ and the second is the region $\iota_{\var{base},\var{end}}^0$
%\lau{doesn't this invariant allow the newly allocated memory to contain capabilities?} A: It does not capture this. One possiblity is to make new standard all zero invariant.
 for an appropriately-sized interval $\var{base},\var{end}$ and the new footprint part satisfies this new world $\heapSat[\hsfoot']{n}{W'}$. 
%\lau{at what step should this hold? Just some step $n'<n$?} A: try with n, if that does not work, then say something like: there exists j such that malloc terminates in j steps and use n-j.
%\lau{It should probably be more precise what it means to ``invoke'' $\malloc$ and for $\malloc$ to ``return'' something.} A: yes. If we won't abstract notion of this, the standard way to formulate it would be with Hoare triples.
%\lau{How do we express that $\malloc$ can be ``trusted'' in the sense that it won't save any capabilities like one to the newly allocated memory or the return capability.} A:  it allows for read capabilities! To capture this, we need either non-interference or some kind of trace semantics!

@@

If we continue running from $\Phi'$ until just after the point we have executed $\jmp{\adv}$, then  $\pcreg$ points to an execute capability for the adversary, and  we reach a configuration $\Phi''$ which is the same as $\Phi$ except:
\begin{itemize}
\item $\memheap[\Phi''](0)$ contains $-1$
\item capabilities for \emph{the counter} and io are saved to the designated addresses in the tdtp
\item the ticket dispenser code resides on the heap starting at the address $\var{base}$ (the code is as seen in Figure~\ref{tab:tick-disp-heap})
\item $\memreg[\Phi''](\pcreg) = (\exec,\advb,\adve,\advb)$
\item $\memreg[\Phi''](r_0) = (\entry,\initb,\inite,\reta)$
\item $\memreg[\Phi''](r_1) = (\entry,\var{base},\var{end},\var{base})$
\item all other registers contain 0
\end{itemize}
We now want to use the FTLR. To do so, we need to define a world that represents our current heap. We pick the following world:
\begin{itemize}
\item $\iota_{\var{io}}$ the region governs the heap segment that only has address $0$. The value of address $0$ should be $-1$ or $\geq 0$ and even.  % The region that governs that output address
\item $\iota_{\var{td}}$ this region governs heap segments with the domain $[\var{base},\var{end}]$ for the addresses $[\var{base},\var{end} - 1]$ they are the same as in $\Phi''$. $\hs(\var{base})$ is positive and even.
\item $\iota_{\var{tdtp}}$ this region governs the ticket dispenser test programs. The domain of a heap segment in this region is that of the ticket dispenser test program and the contents of the heap segment should be the same as it is in $\Phi''$.
\item $\iota_{\advb,\adve}$
\item $\iota_{\malloc}$
\end{itemize}
To use the FTLR, we have to argue that the read condition is satisfied for the region, the adversary is in. This is easily shown, as we have chosen the region for the adversary to be the standard one, so with $\advb$ and $\adve$ as witnesses, we need to show $\iota{\advb,\adve} \subseteq \iota{\advb,\adve}$.

>>>>

In the JS paper, the ticket dispenser program dereferences the counter and returns the value. It does not seem like a similar thing would work here as the adversary can do even more. In particular, the adversary can execute a $\halt$ instruction that will cause the machine to halt (successfully). The adversary could also choose never to return to the return address that we set up for him.
\begin{lemma}[Ticket dispenser test]\dominique{I would prefer more precise, symbolic statements of assumptions and less prose.}
\label{lem:tckt-disp}
 Given a configuration $\Phi$, if
 \begin{description}
 \item[$\boldsymbol \memheap$] contains the ticket dispenser test program starting at the address labelled $\initb$, 
 \item[$\boldsymbol \memheap$] contains the adversary code starting at label $\advb$. The adversary is assumed to have no capabilities on the heap, but it has access to the ambient $\malloc$ capability! \lau{maybe mention that malloc is also on the heap}
 \item[$\boldsymbol \memheap$] contains the malloc code, and the malloc code satisfy $\iota_{\malloc}$, i.e., $\forall n \ldotp \heapSat[{\heap\restrictfun{\{\malloc_{\var{base}}, \dots ,\malloc_{\var{end}}\}}}]{n}{W_{\malloc}}$.
 \item[All the above programs are disjoint.]
 \item[$\boldsymbol{\adv}$ and $\boldsymbol{ \codelabel{malloc}}$] capabilities are ambiently available at all times.
 \item[$\boldsymbol{ \memreg(\pcreg)}$] is $(\rwx,\initb,\inite,\initb)$.
 \item[$\boldsymbol{ \memreg(\rio)}$] is $(\readwrite,0,0,0)$.
 \item[the remaining registers] all contains 0,
 \end{description}
 and $\Phi \rightarrow^* (\halted,\Phi')$, then the I/O address (that is $\memheap[\Phi'](0)$) contains either $-1$ or a number that is $\geq 0$ and even.
\end{lemma}
In the initial configuration the $\pcreg$ register is assumed to have an execute and write permission for the area of the heap where the program resides.

The ticket dispenser test program uses $\malloc$. To be able to reason about programs that use $\malloc$, we assume that it follows the following specification. If you provide it with one argument, namely the size of the allocation you wish and a return capability, then at some point $\pcreg$ will contain an executable capability based on this return capability and the return register will contain a $\rwx$ capability for the freshly allocated memory. 
\begin{specification}[Malloc v.2]
  \begin{align*}
    &\forall \Phi \in \Mems \ldotp \forall \hsfoot, \hsframe \in \HeapSegments\ldotp\\
    &\quad \forall n, i, \size \in \nats\ldotp\forall b,e,a,m_b,m_e\in \Addrs\ldotp\forall p \in \Perms \ldotp\\
    &\qquad W_{\malloc} = [i \mapsto \iota_{\malloc}] \land \\
    &\qquad \dom(\hsfoot) = [m_b,m_e] \land \\
    &\qquad \memheap = \hsfoot \uplus \hsframe \land\\
    &\qquad \heapSat[\hsfoot]{n}{W_{\malloc}} \land \\
    &\qquad \memreg(r_1) = \size \land \\
    &\qquad \size \geq 0 \land \\
    &\qquad \memreg(r_0) = (p,b,e,a) \land \\
    &\qquad \memreg(\pcreg) = (\entry,m_b,m_e,m_b) \\
    &\qquad\Rightarrow \\
    &\qquad \quad\exists \Phi' \in \Mems \ldotp \exists \hsfoot', \hsframe' \in \HeapSegments\ldotp\\
    &\qquad \qquad \exists j,k \in \nats \ldotp \exists b',e'\in \Addrs \ldotp \\
    &\qquad \qquad \quad \Phi \step[j]\Phi' \land \\
    &\qquad \qquad \quad \memheap[\Phi']=\hsfoot' \uplus \hsframe' \land\\
    &\qquad \qquad \quad W' = [i \mapsto \iota_{\malloc}, k \mapsto \iota^0_{b',e'}] \land \\
    &\qquad \qquad \quad \heapSat[\hsfoot']{n-j}{W'} \land \\
    &\qquad \qquad \quad \memreg[\Phi'](\pcreg) = (\updatePcPerm{p},b,e,a) \land \\
    &\qquad \qquad \quad \memreg[\Phi'](r_1) = (\rwx,b',e',b') \land \\
    &\qquad \qquad \quad \forall r \in \RegName \setminus \{\pcreg,r_1\} \ldotp \memreg[\Phi'](r) = 0
\end{align*}
\end{specification}
\lau{I am not satisfied with the pc in the premise.}

\begin{specification}[Malloc]\dominique{can you state this more systematically, with less prose, more explicit assumptions and conclusion?}
Take $W_{\malloc}$ the world with exactly one island, namely the unspecified island $\iota_{\malloc}$ that governs the internal state of $\malloc$.  If we invoke $\malloc$ in a memory $h = \hsframe \uplus \hsfoot$
%\lau{Is $\hsframe$ supposed to be the rest of memory, so $h$ is something we can execute in? Is $\malloc$ allowed to allocate more memory, i.e., how does $\hsfoot$ relate to $\hsfoot'$?} A: The frame contains anything that has been allocated and the footprint contains the rest. That is the footprint contains mallocs internal state as well as the free memory.
% Notice: the current worlds do not support malloc allocating memory as the invariants cannot evolve.
that is the disjoint sum of a frame part and a footprint part and the latter satisfies this world up to steps $n$, i.e., $\heapSat[\hsfoot]{n}{W_{\malloc}}$. Then $\malloc$ will successfully terminate and return a result in a heap $h' = \hsframe \uplus \hsfoot'$ that is the disjoint sum of the same unmodified frame part and a potentially new footprint part, such that there is a future world $W'$ of $W_{\malloc}$ consisting of exactly two islands, where the first is an instance of $\iota_{\malloc}$ and the second is the region $\iota_{\var{base},\var{end}}^0$
%\lau{doesn't this invariant allow the newly allocated memory to contain capabilities?} A: It does not capture this. One possiblity is to make new standard all zero invariant.
 for an appropriately-sized interval $\var{base},\var{end}$ and the new footprint part satisfies this new world $\heapSat[\hsfoot']{n}{W'}$. 
%\lau{at what step should this hold? Just some step $n'<n$?} A: try with n, if that does not work, then say something like: there exists j such that malloc terminates in j steps and use n-j.
%\lau{It should probably be more precise what it means to ``invoke'' $\malloc$ and for $\malloc$ to ``return'' something.} A: yes. If we won't abstract notion of this, the standard way to formulate it would be with Hoare triples.
%\lau{How do we express that $\malloc$ can be ``trusted'' in the sense that it won't save any capabilities like one to the newly allocated memory or the return capability.} A:  it allows for read capabilities! To capture this, we need either non-interference or some kind of trace semantics!

@@

If we continue running from $\Phi'$ until just after the point we have executed $\jmp{\adv}$, then  $\pcreg$ points to an execute capability for the adversary, and  we reach a configuration $\Phi''$ which is the same as $\Phi$ except:
\begin{itemize}
\item $\memheap[\Phi''](0)$ contains $-1$
\item capabilities for \emph{the counter} and io are saved to the designated addresses in the tdtp
\item the ticket dispenser code resides on the heap starting at the address $\var{base}$ (the code is as seen in Figure~\ref{tab:tick-disp-heap})
\item $\memreg[\Phi''](\pcreg) = (\exec,\advb,\adve,\advb)$
\item $\memreg[\Phi''](r_0) = (\entry,\initb,\inite,\reta)$
\item $\memreg[\Phi''](r_1) = (\entry,\var{base},\var{end},\var{base})$
\item all other registers contain 0
\end{itemize}
We now want to use the FTLR. To do so, we need to define a world that represents our current heap. We pick the following world:
\begin{itemize}
\item $\iota_{\var{io}}$ the region governs the heap segment that only has address $0$. The value of address $0$ should be $-1$ or $\geq 0$ and even.  % The region that governs that output address
\item $\iota_{\var{td}}$ this region governs heap segments with the domain $[\var{base},\var{end}]$ for the addresses $[\var{base},\var{end} - 1]$ they are the same as in $\Phi''$. $\hs(\var{base})$ is positive and even.
\item $\iota_{\var{tdtp}}$ this region governs the ticket dispenser test programs. The domain of a heap segment in this region is that of the ticket dispenser test program and the contents of the heap segment should be the same as it is in $\Phi''$.
\item $\iota_{\advb,\adve}$
\item $\iota_{\malloc}$
\end{itemize}
To use the FTLR, we have to argue that the read condition is satisfied for the region, the adversary is in. This is easily shown, as we have chosen the region for the adversary to be the standard one, so with $\advb$ and $\adve$ as witnesses, we need to show $\iota_{\advb,\adve} \subseteq \iota_{\advb,\adve}$.
